Stack Safety as a Security Property

### 2020-07-05 main.tex
Wrote some future work, spun out examples into a separate file for readability

The call stack is an increasingly\apt{``a perennially'' ?} popular target for attacks
exploiting memory safety vulnerabilities, with potential consequences
ranging from attackers reading or corrupting private stack data, to
highjacking a program's control flow. To prevent such attacks, many
techniques have been proposed in the literature\leo{cite, point out?
  stack canaries, cherry, nicks...}\bcp{I think pointing out makes a
  stronger point.} aiming to achieve ``stack
  safety''.

But what {\em is} this stack safety of which we speak? A precise formal
definition is lacking in the literature, with the notion usually being
defined negatively: through examples of stack unsafety
\rb{usually though not always; as noted below there are other
  approaches}\bcp{Yes, we have to be very careful about this}. Two classical
examples of stack unsafety appear in Figure~\ref{fig:stackunsafety}.
\apt{is this gone?}
\leo{Goal: on the left, a bad read; on the right, a bad write.}
\leo{Do we also need an example of ROP/control flow for the new
  part?}\bcp{If it is not too complicated to explain, we should definitely
  put it
  here.  If it is complicated, we can just say in text that there are other
  sorts of attacks too, and that we deal with them, and give a forward
  pointer.}
Recently, a \leo{inspiring/awesome/something good}\bcp{Maybe just say that
  we are only aware of one, quite recent, attempt to do something better} work reasoned
about a stack safety enforcement mechanism using capabilities\leo{cite};
but even here, stack safety is defined with respect to an abstract
machine that ``satisfies stack safety by construction,'' begging
the question of what {\em is} the property that holds by construction?
\apt{Suggest to move first sentence of next para up to this one.}

\leo{StackTokens} provides some intuition towards answering that question,
by informally describing two components that comprise stack safety:
local state encapsulation and well-bracketed control flow. The first
contribution of the present paper is to
turn this intuition into two formal \leo{extensional}\bcp{Not sure we can
  claim this} security properties,
that together form a formal definition of stack safety.
\rb{I think the merit of StkTokens goes beyond mere intuition. There is a lot of
rigor in it, even if some notions are implicit or not fully
developed.}\bcp{Yes, and let's remember that they are highly likely to be
reviewing this paper!  Let's be as generous as we can be without
under-selling our delta.}

The underlying notion behind local state encapsulation is the
intention to protect callers from callees: a callee should be not be
able to access or overwrite a caller's private data. We formalize this
intuition as a security property with both confidentiality and
integrity components.
%
Intuitively, {\em stack confidentiality} promises that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g. a caller's private data or uninitialized memory).  {\em Stack
  integrity} guarantees that, if a function call ever returns, the caller's
stack data at the point of return will be identical to what it was before
the call.  \bcp{We need to point out somewhere that we are only dealing with
stack safety, though we discuss (somewhere) how our property can be combined
with the heap safety property of [Azevedo et al]}

While integrity and confidentiality are relatively standard security
properties, the stack-safety setting raises an interesting additional
challenge:
the notion of who the attacker is shifts throught execution. Not only
does a caller need to be protected from its callee, but the callee
itself needs to be proteced from {\em its} callees.
We address this challenge by varying the observation power\apt{is this
  the right word? doesn't it conflict with our later use of ``observation''?}
of the attacker at call points, and demonstrate that our policy is
implementable by pre-existing monitoring mechanisms. \bcp{The two parts of
  the sentence are about two very different things.  The second deserves its
  own paragraph, IMO.}\apt{+1}

Similarly, the underlying notion behind well-bracketed control flow is
the need to protect the call structure itself: a callee should not be
allowed to jump to the middle of an arbitrary block of code either
directly (via a jump instruction) or indirectly (by corrupting stack
data and returning). Well-bracketed control flow establishes safety between
functions: all code is uniquely owned by some function, and all jumps between functions
are either calls that enter at a marked entry point or returns that enter
just after their associated call in the standard nested call structure.

%\leo{There needs to be some informal description
%  of the wbcf part here. Sean?}

In summary, we offer the following contributions\apt{is this up to date with current paper organization?}:
\begin{itemize}
\item \leo{More? Extensional}\bcp{We might need to invent a word for this---''quasi-extensional''?} formal characterizations of the
  intuitive concepts of local state encapsulation and
  well-bracketed control flow~(\Cref{sec:lse-and-wbcf}).  As evidence that this
  property is strong enough to capture the folk intuition, we show
  that the stack unsafe examples from the literature also fail our
  property~(\Cref{sec:disc}).\apt{Something about relation to StkTokens formalization?}
\item As evidence that the property is not too strong\apt{But we add the lazy version because the strict version \emph{is} too strong, in some sense.}, we show that it is
  realizible by
  a previously proposed micropolicy enforcement
  mechanism~(\Cref{sec:micropolicy}). \leo{And it says yes on a
    substantial body of code}. We also present a variation of the
  property + policy that is lazy, easier/more efficient to implement,
  and provably weaker (\Cref{sec:lazy}).
\item A provably stronger (inductive? step-by-step/stepwise) variant
  of this property tuned for testing/verification.  We test the whole
  thing with a cool framework. \bcp{Which section?}
\end{itemize}
Section~\ref{sec:relwork} discusses related work, and
Section~\ref{sec:future} sketches directions for future work.
%
We begin with a running example that we're going to keep coming back
to throughout the paper (Section~\ref{sec:running-example}) and some
technical preliminaries (Section~\ref{sec:prelim}).

\section{Stack Safety and Unsafety by Example}
\label{sec:running-example}

@@

\subsection{Running Example}

@@

\bcp{Notation nit: in many assembly notations, square brackets mean
  ``{\em indirect} through this location...''  And a very small nit: Change
  $\rsp-2$ to $\rsp\mathord{-}2$}

@@

\section{Discussion}
\bcp{Make this a subsection?}
\label{sec:disc}

  \paragraph{Examples}

\bcp{Belongs with the running example:}
  \bcp{Illustrate the definition by using it to discuss a range of ``stack
    unsafety'' examples from the literature. }

  \paragraph{What about tail recursion and other non-stack-like control flow}

\bcp{And this and the next belong in Future Work...}

    While these ideas are here presented in a stack-centric style, they could
    be extended to heap safety in a system extended with a heap. Heap safety,
    like stack safety, may be framed as a form of noninterference
    [cite Meaning of Memory Safety]. For each certain component marked as
    containing a valid pointer to a heap object, we can propagate that
    validity when the pointer moves or is used in pointer arithmetic. Heap
    objects to which a valid pointer is in a register are reachable, and
    treated as low confidentiality and low integrity.

@@

We've assumed that callees always have less access than their callers, but
in real systems things are more complicated (because of objects, static
variables in C, software compartmentalization, coroutines, ...).  What we've
defined is a very ``pure'' variant of stack safety.  In future work, we
should go further. :-)

>>>>

The call stack is a perennially popular target for attacks
exploiting memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
hijacking of the program's control flow. To prevent such attacks, many
software and hardware techniques for stack safety have been proposed, % in the literature and implemented in production systems,
including stack canaries~\citep{Cowan+98},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
split stacks~\citep{Kuznetsov+14},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \apt{Mostly from nick;there could be more}

But what {\em is} this ``stack safety'' of which we speak? Precise formal
definitions are lacking in the literature; instead, the notion of safety
is usually defined negatively, by enumerating the bad things that a safe system
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%with the notion usually being
%defined negatively: through examples of stack-based exploits 
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ... 
We are only aware of one, quite recent, attempt to frame a positive definition.
\citet{Skorstengaard+19} define stack safety as
the conjunction of two properties, local state encapsulation and well-bracketed control flow.
They formalize these properties by defining an abstract capability-based machine   
that ``satisfies [stack safety] by construction.'' We believe there is still room
to craft a more abstract and less intensional characterization of
stack safety, and that is the goal of the present paper. \apt{still needs tuning}

Our approach is to give alternative formal
definitions of local state encapsulation and well-bracketed control flow
using the conceptual framework of language-based security.
%
The underlying notion behind local state encapsulation is the
intention to protect callers from callees: a callee should be not be
able to access or overwrite a caller's private data. We formalize this
intuition as a security property with both confidentiality and
integrity components.
%
Intuitively, {\em stack confidentiality} promises that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g. a caller's private data or uninitialized memory).  {\em Stack
  integrity} guarantees that, if a function call ever returns, the caller's
stack data at the point of return will be identical to what it was before
the call.  \bcp{We need to point out somewhere that we are only dealing with
stack safety, though we discuss (somewhere) how our property can be combined
with the heap safety property of [Azevedo et al]}

% BCP idea: contrast intrinsic vs. extrinsic property definition
% Remember to talk about spatial vs temporal safety

While integrity and confidentiality are standard security
properties, applying them in the stack-safety setting raises a novel challenge:
the notion of ``attacker'' changes during execution. Not only
does a caller need to be protected from its callee, but the callee
itself needs to be protected from {\em its} callees.
We address this challenge by adjusting the integrity and confidentiality
of different stack frames at each call or return transition between functions.\apt{connection to ``attacker'' still vague.}

The underlying notion behind well-bracketed control flow is
the need to protect the call structure itself: a callee should not be
allowed to jump to the middle of an arbitrary block of code either
directly (via a jump instruction) or indirectly (by corrupting stack
data and returning). Well-bracketed control flow establishes safety between
functions: each instruction is uniquely owned by some function, and all jumps between functions
are either calls that enter at a marked entry point or returns that re-enter
just after their associated call in the standard nested call structure.
\apt{clarify what we are contributing here.}

To provide evidence that our definitions capture the inituitive notion of stack safety,
we show that they rule out a wide \apt{really?} range of exploitable bad behaviors.\apt{``from the literature''??}
To assess whether they are implementable, 
we compare them against the proposed capability-based enforcement
mechanism of \citet{Skorstengaard+19} \apt{we really should} and the existing tag-based enforcement mechanism
of \citet{DBLP:conf/sp/RoesslerD18}.

In summary, we offer the following contributions:
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for   this---''quasi-extensional''?}
  Abstract formal properties capturing the 
  intuitive concepts of local state encapsulation and well-bracketed
  control flow~(\Cref{sec:lse-and-wbcf}) in a very general setting. These properties
  rule out common examples of stack memory abuse.
\item
  A case study of the Depth Isolation policy of \citet{DBLP:conf/sp/RoesslerD18}, based on hardware tags~(\Cref{sec:enforcement}).
  This policy enforces our stack safety property, and also a (provably) stronger property that is easier to test.
  \apt{If we have no testing, do we even want to talk about the testing property?}
\item
  A refinement of the case study that considers the more efficient Lazy Tagging and Clearing policy 
  of \citet{DBLP:conf/sp/RoesslerD18}~(\Cref{sec:lazy}).  This policy does not enforce our property, but we give a
  variant of the policy that does enforce a (provably) weaker variant of our property. \apt{Awkward, but the truth I fear.}

  %\item As evidence that the property is not too strong\apt{But we add the lazy version because the strict version \emph{is} too strong, in some sense.}, we show that it is
%  realizible by
%  a previously proposed micropolicy enforcement
%  mechanism~(\Cref{sec:micropolicy}). \leo{And it says yes on a
%    substantial body of code}. We also present a variation of the
%  property + policy that is lazy, easier/more efficient to implement,
%  and provably weaker (\Cref{sec:lazy}).
%\item A provably stronger (inductive? step-by-step/stepwise) variant
%  of this property tuned for testing/verification.  We test the whole
%  thing with a cool framework. \bcp{Which section?}
\end{itemize}
Section~\ref{sec:relwork} discusses related work, and
Section~\ref{sec:future} sketches directions for future work.
%
We begin with a running example that we refer
to throughout the paper (Section~\ref{sec:running-example}) and some
technical preliminaries (Section~\ref{sec:prelim}).

\section{Running Example}
\label{sec:running-example}

@@

@@

@@

@@

  We have presented a pure notion of stack safety that omits many complicating
  factors common to real system. In particular, our model as presented exploits
  the simplifying assumption that privilege never increases during a function
  call, and that therefore a contour computed once on the entry to a function
  describes its privilege throughout. Our future plans are to extend the model
  with common language features, some of which violate this assumption. Here we
  describe the basic principles by which our model can be extended to some of
  these cases.

  \paragraph{Stack-derived Pointers}

    Common programming idioms involve a caller passing a pointer to its local
    data into a callee. Using standard notions of pointer provenance [cite],
    we can extend our model to distinguish safe use of this idiom from true
    stack safety violations. We require additional annotation to identify
    which addresses within a function's stack frame correspond to distinct
    objects. Then a valid pointer to an object is one derived through legal
    arithmetic from its base address, and at each call point, an object's
    addresses are marked low confidentiality and integrity if a valid pointer
    exists in a register or in the transitive closure of the callee's
    accessible memory. All pointers to an object cease to be valid when it is
    deallocated along with its stack frame.

    Heap safety, like stack safety, may be framed as a form of noninterference
    [cite Meaning of Memory Safety]. The above model describing safety of
    shared stack-allocated objects extends naturally to heap safety,
    complicated by the fact that these objects are not deallocated by returns,
    and therefore a function's privilege may increase or decrease after its
    entry point. For instance, if a callee allocates a heap object and returns
    the pointer to its caller, that object's addresses become accessible in
    contradiction to a contour computed at the caller's entry. So such an
    extension must enforce trace confidentiality and integrity properties
    separately on continuous segments of each call, with contours computed at
    each crossing between caller and callee.

  \paragraph{Non-stack control flow}

    Control structures beyond well-bracketed calls and returns require
    modifications to the model. Tail calls, for instance, reuse the caller's
    stack frame for its callee, and every nested tail call returns
    simultaneously to the top non-tail-calling function. Under normal stack
    integrity, the first function in a chain of tail calls returns and violates
    integrity, while subsequent tail calls never actually return at all.
    Instead we need a variant that treats the entry of a tail call as the
    return of its caller for purposes.

    More complex is the addition of a coroutine model, in which multiple stacks
    respect stack safety internally, but might also yield to one another.
    Here we must distinguish three levels of integrity: accessible, in the same
    stack, and in another stack. Stack integrity requires that same-stack data
    be unchanged when a function returns, but other-stack data may change. A
    similar {\em yield integrity} property requires that, from a yield out of
    a coroutine to the next yield back into it, all accessible and same-stack
    data are unchanged. Confidentiality properties are split similarly.

\section{Discussion}

\bcp{Make this a subsection?}
\label{sec:disc}
